Process-based phenotypes: an application to the ontology of cell physiology

### 2011-11-30 paper/drinking.tex

We assume that the classes of an ontology are
satisfiable, i.e. that they have instances

Classes in an ontology may be unsatisfiable: a class $C$ is
unsatisfiable, if it is not possible that the class has
instances. Unsatisfiable classes are commonly the result of a
contradictory class definition. That a class is unsatisfiable is often
a special kind of unintended consequence that can be drawn from an
ontology.

@@

Processes commonly have participants\footnote{It may be argued that
  processes necessarily have participants or that processes may have
  no participants. However, these discussions do not affect our aims
  here and take no position in this regard.}. Participants of
processes are entities that are present at time points (i.e., they
have no temporal parts) and may undergo changes during the process
\cite{Herre2006, Herre2010}.  We can further distinguish different
{\em roles} of participation in a process \cite{Loebe2007}. For
example, a runner may participate in a {\em Triathlon} process as a
{\em Runner (role)}, while another person can participate in the same
process but in a different role (such as {\em Referee}).

Amongst the participants of some processes, we will distinguish
between {\em inputs} and {\em outputs} of processes. Chemical
reactions, for example, are processes with definite inputs (the {\em
  reactants}) and outputs ({\em the products}).

@@

\section{Results}
\subsection{Attributes of processes}
We develop a model of process attributes that is applicable for
representations of physiology and related phenotypes. In principle, we
distinguish between three different kinds of process attributes:
the first are process attributes that arise directly from processes
and include {\em duration} and {\em temporal location}; the second are
attributes that arise from processes and their parts and include {\em
  frequency}; and the third are attributes that arise from processes
and qualities of their participants, and include {\em flow rates}.

Attribute that can be directly linked to a process arise from
processes' temporal extension. For example, a duration is an attribute
that characterizes the temporal extent of a process and is similar to
{\em Length}, {\em Area} and {\em Volume} for one-, two- and
three-dimensional physical objects. A {\em Temporal location}
positions the time interval at which the process occurs with respect
to a reference coordinate system.

@@

However, not all rates of a process depend on attributes of the
process participants. In particular, a {\em frequency of occurrence}
or {\em event rate} refers to the number of occurrences of a type of
process during a reference process. For example, a {\em rate of heart
  beating} refers to the number of {\em Heart beating} processes that
occur within a reference process (e.g., a process in which the heart
participates with a duration of one minute). Further attributes that
depend on types of processes with regard to a reference process are
{\em distribution patterns}, i.e., how the occurrences of processes of
a particular type are distributed within a reference process. For
example, the heart may beat {\em rhythmically} or {\em arrhythmically}
within a period of time.

@@

While some processes can be subdivided indefinitely without changing
the type of process, others cannot. A class $C$ of {\em continuous
  processes} is a class that has processes $p$ as instance such that
all temporal parts of $p$ are also instances of $C$. Examples include
{\em continuous movements} or {\em mass flow} processes. On the other
hand, {\em discrete processes} can be subdivided into stages of
activity and stages of inactivity: a class $C$ of {\em discrete
  processes} is a class with processes $p$ as instance and there is a
part of $p$ that is not an instance of $C$. An example is {\em heart
  beating} which has periods of activity (a single heart beat) and
inactivity.

We may further attribute a {\em frequency} or {\em rate} to an object
instead of a process. For example, a heart that beats {\em now} with a
frequency of 80 beats per minute, or a car that is moving at a speed
of 180 kilometres per hour {\em at a particular point in time} (e.g.,
as observed with a speed camera) can be considered attributes of the
objects (the heart or the car), not attributes of the processes in
which the objects participate. However, these are {\em different}
kinds of attributes. Rates, when considered as attributes of objects,
may be explicitly defined using rates of processes. For example, the
heart beating frequency of a particular heart $h$ at a time point $t$
is the frequency of a reference heart beating process in which $h$
participates. Such a reference process is necessary in order to obtain
a value for a frequency even when no {\em heart beating} process is
occurring. However, the frequency is only an attribute of the heart in
virtue of such a reference process in which {\em heart beating} is
actually occurring.  This reference process can be uniquely determined
for continuous processes (such that the frequency or rate of an object
at a time $t$ is the frequency or rate of the infinitesimally small
process around $t$), and is ambiguous for discrete processes (in which
case the reference process must be explicitly stated).

Figure \ref{fig:patterns} illustrates some examples of
non-comparative and comparative process attributes.

>>>>

Automated reasoning can further determine whether classes in an
ontology are unsatisfiable: a class $C$ is unsatisfiable, if it is
impossible for the class to have any instances. Unsatisfiable classes
in an ontology are commonly the result of a contradictory class
definition. % That a class is unsatisfiable is often a special kind of
% unintended consequence that can be drawn from an ontology.

@@

Processes commonly have physical objects as participants. Physical
objects are entities that are present as a whole at time points (i.e.,
they have no temporal parts) and may persist through time, i.e., they
may undergo changes during the process \cite{Herre2006, Herre2010}.
We can further distinguish different {\em roles} of participation in a
process \cite{Loebe2007}. For example, a runner may participate in a
{\em Triathlon} process as a {\em Runner (role)}, while another person
can participate in the same process but in a different role (such as
{\em Referee}).

% Amongst the participants of some processes, we will distinguish
% between {\em inputs} and {\em outputs} of processes. Chemical
% reactions, for example, are processes with definite inputs (the {\em
%   reactants}) and outputs ({\em the products}).

@@

\section{Results}
\subsection{Attributes of processes}
We develop a model of process attributes that is applicable for
representations of physiology and related phenotypes. In principle, we
distinguish between three different kinds of process attributes: the
first are process attributes that arise directly from processes and
include {\em duration} and {\em temporal location}; the second are
attributes that arise from processes and their temporal parts and
include {\em frequency} and {\em onset}; and the third are attributes
that arise from processes and qualities of their participants, and
include {\em flow rates}.

Attributes that can be directly linked to a process arise from
processes' temporal extension. For example, a duration is an attribute
that characterizes the temporal extent of a process and is similar to
{\em Length}, {\em Area} and {\em Volume} for one-, two- and
three-dimensional physical objects. A {\em Temporal location}
positions the time interval at which the process occurs with respect
to a reference coordinate system.

@@

However, not all rates of a process depend on attributes of the
process participants. In particular, a {\em frequency of occurrence}
or {\em event rate} refers to the number of occurrences of a type of
process during a reference process. For example, a {\em rate of heart
  beating} refers to the number of {\em Heart beating} processes that
occur within a reference process (e.g., a process in which the heart
participates with a duration of one minute). Further attributes that
depend on types of processes with regard to a reference process are
{\em distribution patterns}, i.e., how the occurrences of processes of
a particular type are distributed within a reference process. For
example, the heart may beat {\em rhythmically} or {\em arrhythmically}
within a period of time (see Figure \ref{fig:patterns}).

@@

While some processes can be subdivided indefinitely while retaining
certain kinds of attributes, others cannot.
% A class $C$ of {\em continuous
%   processes} is a class that has processes $p$ as instance such that
% all temporal parts of $p$ are also instances of $C$. 
Examples of processes that can be divided include {\em continuous
  movements} or {\em mass flow} processes, for which all parts have a
{\em speed} or {\em flow rate} attribute. On the other hand, some
processes can be subdivided into stages of activity and stages of
inactivity and cannot arbitrarily be divided. For example, a process
of {\em heart beating} has periods of activity (a single heart beat)
and inactivity. Consequently, not all parts of the process have a {\em
  heart rate} attribute.

We may further attribute a {\em frequency} or {\em rate} to an object
instead of a process. For example, a heart that beats {\em now} with a
frequency of 80 beats per minute, or a car that is moving at a speed
of 180 kilometres per hour {\em at a particular point in time} (e.g.,
as observed with a speed camera) can be considered attributes of the
objects (the heart or the car), not attributes of the processes in
which the objects participate. However, these are {\em different}
kinds of attributes. Rates, when considered as attributes of objects,
may be explicitly defined using rates of processes. For example, the
heart beating frequency of a particular heart $h$ at a time point $t$
is the frequency of a reference heart beating process in which $h$
participates. Such a reference process is necessary in order to obtain
a value for a frequency even when no {\em heart beating} process is
occurring. However, the frequency is only an attribute of the heart in
virtue of such a reference process in which {\em heart beating} is
actually occurring.  This reference process can be uniquely determined
for processes such as {\em continuous movement}, where the rate of an
object at a time $t$ is the rate of the infinitesimally small process
that occurs around $t$. The reference process is ambiguous for
processes such as {\em heart beating}, and the reference process must
be explicitly stated.

% Figure \ref{fig:patterns} illustrates some examples of
% non-comparative and comparative process attributes.
